## Visualizing with ggplot2

## Preparation
# Masked function names

Since we will be using dplyr extensively throughout this session, let’s go ahead and attach it to our local environment via the library() function.


```{r}
library(dplyr)
```

# Load Dataset
Lets continue by loading in the dataset we were working with during Week # 4.


```{r}
StorageLocation <- file.path("data/")
```


```{r}
TheCSV <- list.files(StorageLocation, pattern = ".csv", full.names = TRUE)
Data <- read.csv(TheCSV, check.names=FALSE)
```


```{r}
head(Data, 3)
```

# Identifying column value types
And let’s quickly evaluate the columns to determine what type of values each contains


```{r}
str(Data)
```

The type of values that a column contains are particularly important to be aware of when working with ggplot2 plots. For example, we can see that several columns contain character values:


```{r}
Data |> select(bid, Condition, Date, infant_sex, ptype) |> head(6) |> str()
```

While spotting the character columns from the str() output, and isolating them using select() works, we could achieve something similar using the is.character() function in combination with the where() function, which the dplyr package actively borrows (ie. exports) from the tidyselect package.

```{r}
Data |> select(where(is.character)) |> head(6) |> str()
```

We can switch out is.character() to identify columns containing other value types

```{r}
Data |> select(where(is.numeric)) |> head(6) |> str()
```

```{r}
Data |> select(where(is.logical)) |> head(6) |> str()
```

# Reformatting Column Value Types

ooking at the str() output, we can spot that the Date column is currently showing as a character value:

```{r}
Data |> select(Date) |> str()
```

Character values when plotting are generally treated as categorical factors. When plotted, these will arrange according to alphabetical order. In this particular case, this may not be an issue:

```{r}
Data |> pull(Date) |> sort() |> unique()
```

However, if year/month/day is formatted differently, as character values, the alphabetical reordering could result in your dates being scrambled

```{r}
AlternateFormat <- c("26-07-2025", "27-08-2019", "09-09-2025", "16-03-2026")
AlternateFormat |> sort() |> unique()
```

Because of this, I generally recommend reformatting these character type values over to Date type values. I will generally use the the tidyverse lubridate package, as it has various functions that can handle the date format variation. In this case, since the character values appear year-month-day, we can use the ymd() function to update the column type

```{r}
library(lubridate)
```
```{r}
Data$Date <- ymd(Data$Date)
str(Data [, 1:5])
```

You can notice, str() now returns the column type as “Date”.

# Factors
For columns with numeric type values, we previously discussed the difference between integer (ie. whole) and double (containing decimal point) values. When it comes to plotting using ggplot, numeric values will generally be treated as continuous values unless we specify they should be treated as categorical values for plotting.

For example, if we look at the Timepoint column, we can see its numeric values are discrete, corresponding to the timepoint in months when the blood sample was collected (0, 4, 9)

```{r}
Data |> pull(timepoint) |> unique()
```

f we want to avoid having ggplot treating these values as continuous, and instead see them as categorical, we will need to specify to R to treat these as such. We can do this by converting them into factos, using the factor() function.

```{r}
Data1 <- Data
Data1$timepoint <- factor(Data1$timepoint)
str(Data1 [,1:3])
```

# ggplot2
Glancing at Data, let’s try creating a boxplot, with the different timepoints on the x-axis, and the frequency of T cells in the CD45+ gate on the y-axis. We can envision we want something at the end that resembles the following:

Let’s go ahead and actually convert timepoint over to a categorical factor for this example, and mutate the frequency column.

```{r}
Data <- Data |> mutate(TcellProportion=Tcells_count/CD45_count) |> mutate(TcellFrequency=TcellProportion *100) |> mutate(TcellFrequency=round(TcellFrequency, 1))

Data$timepoint <- factor(Data$timepoint)
```

Lets start by going ahead and calling library() for ggplot2

```{r}
library(ggplot2)
```

# Data

```{r}
Plot <- ggplot(Data)
Plot
```

As you can see, nothing is yet outputted at this stage. Let’s however peak behind the curtain.

```{r}
str(Plot)
```

In this first chunk, we see that the data slot of the ggplot object now contains the underlying data.frame object

If we scroll past, we can spot additional slots, which are currently empty, waiting to be filled. We will sporadically check back and see how this changes.

# Aesthetics
Having established the first layer of our plot (Data), we can now start by specifying the next layer, aesthetics. In our case, we were interested in placing timepoint on the x-axis, and placing the TcellFrequency on the y-axis.

```{r}
Plot <- ggplot(Data) + aes(x=timepoint, y=TcellFrequency)
Plot
```

As you can see, additional elements were added to the plot, namely we are now seeing on the axis we specified.

If we glanced behind the curtains, we now see the mapping slot has now been filled in


```{r}
str(Plot)
```

In our case, we are building this plot with each layer being connected by a “+”” argument. However, you will often see both the Data and Aesthetics layers combined together in actual practice. This would look like the following

```{r}
ggplot(Data, aes(x=timepoint, y=TcellFrequency))
```

# No quotation Marks
You may have noticed, we relied on tidyverse to figure out that timepoint and TcellFrequency were column names, and not objects in our environment (for reasons detailed during Week 04). What would have happened if we had included “” around each?

```{r}
ggplot(Data) + aes(x="timepoint", y="TcellFrequency")
```

Yikes! Not what we were expecting.

# Geometries
Having specified our Data and Aesthetics layers, we now reach the important point in our plotting step, deciding on what geometries we want to use. Geometry is the layer that determines how our data is displayed/plotted.

For example, we could use the geom_point() function to set the plot geometry as follows

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_point()
```

We could swap geom_point() out for geom_boxplot()

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_boxplot()
```

or geom_violin()

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_violin()
```

We could also add two geometries

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_boxplot() + geom_point()
```

As you can see, many different geom_ functions exist in both ggplot2, as well as other R packages that expand the plotting capacity further. One I like to use for my own plots is the ggbeeswarm package, specifically the geom_beeswarm() function in conjunction with the geom_boxplots().

```{r}
install.packages("ggbeeswarm")
library(ggbeeswarm)
Plot <- ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_boxplot() + geom_beeswarm()
Plot
```

We will continue building offf this plot version as we go forward.

# if left as continuous
In the case of the above dataset, we had converted timepoint to a factor. But what would our plot have looked like if we had left it in the original continuous format?

```{r}
AlternateData <- read.csv(TheCSV, check.names=FALSE)
AlternateData <- AlternateData |> 
    mutate(TcellProportion=Tcells_count/CD45_count) |>
    mutate(TcellFrequency=TcellProportion *100) |>
    mutate(TcellFrequency=round(TcellFrequency, 1))
ggplot(AlternateData) + aes(x=timepoint, y=TcellFrequency) + geom_beeswarm()
```

As you can see, in this case, the x-axis instead of being spaced categorically, now appears as continuous, with the individual dots being plotted at 0, 4, 9 respectively.

Given that geom_boxplot is expecting a categorical factor, this fails to plot correctly when added on

```{r}
ggplot(AlternateData) + aes(x=timepoint, y=TcellFrequency) + geom_boxplot() + geom_beeswarm()
```

Fortunately, the warning message provides some context we can use when troubleshooting. If we implement the suggestion, we are able to rescue geom_boxplot() plotting attempt even in abscence of specifying timepoint as categorical.

```{r}
ggplot(AlternateData) + aes(x=timepoint, y=TcellFrequency, group=timepoint) + geom_boxplot() + geom_beeswarm()
```

This example generally highlights the importance of being aware of what types of values you are trying to plot, warning messages, and some of the quiky odd-looking plots that can arise when the assumptions are not met.

# Mismatched Assumptions
In this case, the two “geom_” functions we are using are using x-axis value as categorical, and y-axis as a continuous variable. What happens if we used a “geom_” function that has different specifications?

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_histogram() 
```

In the case of the above, geom_histogram() was expecting only one column being present in the aes() argument.

```{r}
ggplot(Data) + aes(x=TcellFrequency) + geom_histogram() 
```

This would similarly be the case for geom_density()

```{r}
ggplot(Data) + aes(x=TcellFrequency) + geom_density() 
```

# Geomeetry Arguments

Lets return to our current beeswarm boxplot, which is currently using only the default geom_boxplot() and geom_beeswarm() functions when building the geometry layer

```{r}
Plot <- ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + geom_boxplot() + geom_beeswarm()
Plot
```

If we think back to what we have been learning about grammar of graphics, each layer is manipulated individually, with subsequent layers combining to make the final plots that we might be interested in. Consequently, if we are making any large changes to the geometry layer, it would be occuring as an argument provided to the geometry determining functions. Lets check to see what our argument options are for both of the above

```{r}
?geom_boxplot
```

```{r}
?geom_beeswarm
```

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=5)
```

Waaaaay toooo large. Let’s tone it back

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5)
```

Better, but now the individual dots being too tightly wrapped together is limiting our capacity to visualize individuals. Lets next attempt to get the dots to spread out more horizontally from each other. Taking a look at the documentation, it looks like “cex” is the argument that might work.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=5)
```

Upps! Too far! Let’s adjust it again

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5)
```

Good enough for now!

# Scales
The next ggplot2 layer we will work is scales. Working with our individual dots, we were able to change the size within the geometry layer. However, we can also modify their appearance at the scale level. Refering back to our original data, we can see there is a metadata column corresponding to whether the PBMCs were from male or female donors.

```{r}
Data |> pull(infant_sex) |> unique()
```

Let’s go ahead and attempt to alter the dots to reflect based on this additional metadata variable. First, we can assign values based on the categorical values that are present for that column

```{r}
shape_sex <- c("Female" = 22, "Male" = 21)

fill_sex <- c("Female" = "red", "Male" = "blue")
```

The two scale layer functions we will use will be scale_shape_manual() and scale_fill_manual().

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=2.5) +
    scale_shape_manual(shape=shape_sex) + scale_fill_manual(fill=fill_sex)
```

Error in `discrete_scale()`:
! unused argument (shape = c(22, 21))
This did not quite work as expected, with our additional arguments not being perceived by the plot. Part of this is due to needing to set up the underlying geometry layer with the necessary arguments as well. After some investigation, lets try

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
    scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex)
```

# Facets
At this point, we have added the Data, Aesthetic, and two Geometry layers. We have a decent working version of our plot object which we can continue to edit as we go forward.

```{r}
Plot <- ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex)
Plot
```

The next layer, Facets, allows creation of separate plots based on an additional variable. For example, if we were interested in seeing the difference between male and female infrants, we could specify the column name within the facet_wrap() argument.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + facet_wrap(~ infant_sex)
```

We could similarly return individual plots for various treatment conditions (Ctrl, PPD, SEB)

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + facet_wrap(~ Condition)
```

We could also use the variant facet_grid() to facet on the basis of two separate variables.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + facet_grid(infant_sex ~ Condition)
```

# Coordinates
Having built out much of the underlying plot, we currently have the following plot

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex)
```

An additional ggplot layer we will occasionally work with is the coordinates layer. It doesn’t change the underlying data, but does the position from which we view with. For example, we can use the coord_cartesian() function to zoom out from the plot shown above, and see the full y-axis values from 0 to 100.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + coord_cartesian(ylim=c(0, 100))
```

# Themes
The final layer in todays exploration of ggplot2 is the “theme” layer. This is where most of the customizations tend to occur when it comes to editing our plots, as most of these are often personal choices. There are a couple ways by which we can implement these, either using a broader “theme_” function, which will set to various defaults, or through the general theme() function, which as long as we look up the correct argument will let us edit pretty much anything.

## Broad Themes
Here are a couple examples of what applying one of the various default “theme_”’s looks like.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + theme_classic()
```

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=2.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + theme_bw()
```

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=2.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) + theme_minimal()
```

## Active Customization
By contrast, instead of (or in addition to) applying a broad theme, we can also modify individual elements of our plots. For example, within theme functions parenthesis, we can silence a particular plot element by setting its theme element equal to the element_blank(). If we were trying to remove the major grid lines, we would run a line of code like this.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) +
theme(
    panel.grid.major = element_blank())
```

If we decided to remove the minor grid lines, we would set it’s equivalent argument equal to element_blank() within the theme (), although we must first add a comma to distinguish it from the previous argument.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) + 
    geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) +
theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())
```

In addition to removing things, we can also modify theme elements. One issue often encountered is the axis-text labels obscure one another. One solution is to rotate them (via the “angle” argument)and horizontally justify their starting position using “hjust” to line up with the original starting position.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) +
theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle=45, hjust=1, size = 16))
```

We can in a similar manner adjust the size of the axis text labels, using the “size” argument.

```{r}
ggplot(Data) + aes(x=timepoint, y=TcellFrequency) +
     geom_boxplot() + geom_beeswarm(size=2.5, cex=3.5, aes(shape=infant_sex, fill=infant_sex)) +
     scale_shape_manual(values=shape_sex) + scale_fill_manual(values=fill_sex) +
theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle=45, hjust=1, size = 12),
    axis.text.y = element_text(size=12))
```

# ggcyto

Let’s return to the initial flow cytometry style plots we generated last time using ggcyto, and see if we can customize them further using what we have learned today.

As we mentioned during Week 5, ggplot2 implemented some major changes going from version 3 to 4. As a result of this, ggcyto had to implement several bug fixes to get the developmental branch back to working order. Consequently, as of the time of this course, you need the following versions (or greater) or both packages to successfully create all the plots.

Let’s begin by loading in the additional R packages we will need

```{r}
library(CytoML)
library(ggcyto)
```

And proceed to load in the FlowJo.wsp example from Week 05 into a GatingSet object.

```{r}
# StorageLocation # Defined Above
FlowJoWsp <- list.files(path = StorageLocation, pattern = ".wsp", full = TRUE)
ws <- open_flowjo_xml(FlowJoWsp)
gs <- flowjo_to_gatingset(ws=ws, name=1, path = StorageLocation, additional.keys = "GROUPNAME")
```

And return to the plot we had left off on:

```{r}
Plot <- ggcyto(gs[6], subset="Tcells", aes(x="CD8", y="CD4")) + geom_hex(bins=100)
```


```{r}
Plot + theme_bw()
```